A Bidirected String Graph Model for Genome Assembly

### 2012-12-10 presentation/paper.tex
paper: network flow

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.6\textwidth]{E_coli.pdf}
		\caption{Transitively-reduced, collapsed bidirected string graph for
		1000 bp perfect reads uniformly sampled every 100 bp from random
		strands of the non-circular version of the 4,639,675 bp genome of {\it
		E. coli} (a common bacterium) using a minimum overlap length of 250 bp.
		This graph was produced by the code that has been
		discussed, but the actual drawing of the picture was done by
		{\it GraphViz}.  The 46,387 reads that were provided as input to the
		assembler were generated by a Perl script.  The bottleneck of the
		algorithm up to this point is \ProgramName{compute-overlaps}, which took
		5 minutes and used a maximum of 984 MiB of memory to find all XXXXXX
		overlaps from these reads, considering XXXXXX pairs of reads that shared
		a 128-mer seed.  The running time and memory usage of
		\ProgramName{compute-overlaps} would need to improved further to handle
		larger genomes.  (Also it still does not handle inexact overlaps, which
		are not considered in this paper.)}
		\label{fig:E_coli}
	\end{center}
\end{figure}

@@

\subsection{Calculating the A-statistic}
\subsection{Mapping contained reads}
\subsection{Network flow}
\subsection{Generalized Eulerian path}

>>>>

A \KeyTerm{cycle} in a bidirected graph is a walk that begins and ends at the
same vertex in such a way that it would be valid to trace the walk again when
the starting vertex is reached at the end of the cycle.

A bidirected graph is \KeyTerm{connected} if the undirected graph produced by
replacing all bidirected edges with undirected edges is connected.  A bidirected
graph is \KeyTerm{strongly connected} if it is connected and for every edge
there is a cycle containing it\cite{Medvedev2009}.

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.6\textwidth]{E_coli.pdf}
		\caption{Transitively-reduced, collapsed bidirected string graph for
		1000 bp perfect reads uniformly sampled every 100 bp from random strands
		of the non-circular version of the 4,639,675 bp genome of {\it E. coli}
		(a common bacterium) using a minimum overlap length of 250 bp.  This
		graph was produced by the code that has been discussed, but the actual
		drawing of the picture was done by {\it GraphViz}.  The 46,387 reads
		that were provided as input to the assembler were generated by a Perl
		script.  The bottleneck of the algorithm up to this point is
		\ProgramName{compute-overlaps}, which took about 10 minutes and used a
		maximum of about 1 GiB of memory to find all 336,227 overlaps from these
		reads, seeding on 128-mers.  The running time and memory usage of
		\ProgramName{compute-overlaps} would need to improved further to handle
		larger genomes.  (Also it still does not handle inexact overlaps, which
		are not considered in this paper.)}
		\label{fig:E_coli}
	\end{center}
\end{figure}

@@

As \Figure{fig:E_coli} shows, the bidirected string graph may still be very
complicated even after transitive reduction and traversal.  In even worse
examples, the graph could easily have thousands of vertices and edges.  We still
need to figure out how to walk through the graph to produce the best assembly.

An \KeyTerm{Eulerian path in a bidirected graph} is a walk that traverses each
bidirected edge exactly one time (only in one of the two possible directions).
Suppose we could find an Eulerian path in the bidirected string graph.  Then
this path would correspond to an assembly that accounts for every overlap and
therefore would be a valid assembly of the reads.  Furthermore, such a path may
allow for vertices to be used for than one time in the assembly to allow for
genomic repeats, which would not be true if we found a Hamiltonian path
instead.\footnote{A Hamiltonian path is also inappropriate here because
unbranched paths have been collapsed into single edges, which {\em must} be
included in the assembly if the reads corresponding to the inner vertices are to
take part in the assembly.  Furthermore, keep in mind that there is not a
one-for-one correspondence between vertices and reads because contained reads
were removed (see \Section{sec:removing_contained_reads)}.}

However, there are multiple problems with the Eulerian path approach.  First,
the graph may not be Eulerian.  In the case that the graph

This implies we really would need to solve the
bidirected equivalent of the Chinese Postman Problem instead.  Even then, the
graph may not be fully connected, which would require that the final walk take
the form of multiple discontinuous walks.  An equally severe problem is that
some overlaps may be spurious, meaning that the two overlapping reads do not
actually form a valid genomic sequence when combined.  Spurious overlaps can be
caused by genomic repeats.  Therefore, we need to allow for the possibility that
a given edge in the bidirected graph is omitted from the assembly if it seems
like a good idea to do so.

The way these problems can be solved is to (1) place a lower bound and an upper
bound on the number of times each edge may be traversed, then (2) formulate and
solve a minimum-cost network flow problem that finds the minimum total traversal
count of edges in the graph in order for every vertex to be balanced (i.e. have
the same indegree as out degree), then (3) find a generalized Eulerian tour
through the graph, traversing each edge the exact number of times that was
calculated in (2) and also allowing for the tour be discontinuous.

\subsection{Calculating the A-statistic}

We would like to use information from the data to restrict the number of times
each edge may be traversed.  This can be done using a calculation called the
\KeyTerm{A-statistic}, or \KeyTerm{arrival-rate statistic}, which is the
logarithm of the probability that an edge represents unique sequence rather than
duplicate sequence.  The A-statistic $A(e)$ of an edge $e$ in a bidirected
string graph is:

\[ A(e) = \frac{n}{G} \Delta - \ln{2} \cdot {k} \]

where $n$ is the total number of reads, $G$ is the estimated genome length,
$\Delta$ is the length of the edge in bp, and $k$ is the number of reads that
support that edge.  To more precise, $k$ is the number of inner vertices in the
unbranched path corresponding to $e$ before the unbranched paths were collapsed.

Edges with a sufficiently high A-statistic are very likely to be single-copy
edges and therefore should be traversed exactly one time.  The remaining edges
may be traversed any number of times (with the exact number being determined
during the network flow calculation), except for edges that originally were
associated with one or more inner vertices, in which case we require the edge be
traversed at least one time so that the reads corresponding to the inner
vertices are used in the reconstruction of the genome (as every read was
presumably sampled from the genome, in the absence of foreign DNA).

Calculation of the A-statistic on each edge in the bidirected string graph is
implemented by the program {\tt calculate-A-statistics}.

\subsection{Mapping contained reads}

\subsection{Minimum-cost network flow}

Formally, a minimum-cost network flow problem on a directed graph takes as input
for each edge $e$ an upper bound $c(e) \ge 0$ on flow and a cost per unit flow
$v(e)$, and for each vertex $v$ a supply (positive) or demand (negative)
$b(v)$ for flow.  It produces as output a flow $x(e)$ on each edge $e$ such
that $\sum_e v(e)x(e)$ is minimized subject to $x(e) \in [0, c(e)]$ and
$\sum_{u \to v}x(u \to v) + b(v) = \sum_{v \to w} x(v \to w)$.\cite{Myers2005}.

We can roughly map our problem onto this definition assigning $c(e) = 0$ when
$e$ is definitely single-copy and $c(e) = \infty$ otherwise, $b(v)$ the sum
of the lower bounds on the traversal counts of all bidirected edges incident to
$v$ with an inward head minus the sum of the lower bounds on the traversal
counts of all edges incident to $v$ with an outward head, and $v(e) = 1$.  The
resulting traversal count $t(e)$ for a bidirected edge $e$ is then the lower
bound on the traversal count for that edge plus $x(e)$, the traversal count
assigned in the network flow calculation.  The traversal count $t(e)$ represents
the total number of times the edge may be traversed in a given assembly,
counting traversals in both directions.

There is, however, a serious complication with this network flow formulation
which was not addressed in \cite{Myers2005}.  We have a bidirected graph, not a
directed graph, so it cannot be trivially mapped to a directed network flow
problem.  A possible workaround for this is to solve directed flow on the
directed equivalent to the bidirected graph, described in
\Subsection{subsec:directed}.  The calculated traversal counts for a directed
edge $e$ between, for example, two vertices $f.E \to g.E$ is not guaranteed to
be the same as the calculated traversal count for the edge that corresponds to
the other direction of the implied bidirected edge, which would be $g.B \to f.B$
in this example.  However, gWe can take the average of the two counts and use this as the
traversal count for the corresponding bidirected edge, and this

%vertices $f.X$ and $g.X

\subsection{Generalized Eulerian path}
